An AI is haunting campus

### 2024-06-18 ai-haunting-campus/ai-haunting-campus.qmd
Created quarto file to render report to pdf

---
output: github_document
bibliography: bibliography/bibliography.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

@@

# An AI is haunting campus^[Karl Marx wrote about a "spectre...haunting Europe--the spectre of communism"; GenAI looks awfully like the shade of the spectre of unregulated surveillance capitalism.]: Some considerations about the use of AI for teaching and learning in post-secondary education

<!-- badges: start -->

<!-- badges: end -->

By: Antonio Páez (paezha@mcmaster.ca)

@@

---

## Contents

- [Background](#background)  
- [The university's mission](#the-mission)  
- [What do AIs do?](#what-do-ais-do)  
- [A note on the terms of the conversation](#terms-of-debate)  
- [But really, what do AIs do?](#really-what-do-they-do)  
- [AI in teaching and learning](#ai-in-teaching-and-learning)  
- [What do GenAIs do for students?](#what-do-ais-do-for-students)  
- [What do GenAIs do to students?](#what-can-ais-do-to-students)  
- [What do GenAIs do for the mega-corporations pushing these tools?](#what-do-ais-to-corporations)  
- [What do GenAIs do to everyone else?](#what-do-ais-do-to-everyone-else)

---

## Background {#background}

@@

## The university's mission {#the-mission}

@@

```{r ai-writing-code, out.width="50%", fig.cap="\\label{fig:ai-bug-fix}Figure 1. An example of Atlassian AI writing code"}
knitr::include_graphics("images/ai-bug-fix.png")
```

@@

```{r ai-litter-box, out.width="50%", fig.cap="\\label{fig:ai-bug-fix}Figure 2. Revolutionizing pet health"}
knitr::include_graphics("images/ai-litter-box.png")
```

@@

@@

```{r plot-simulated-data, message=FALSE, out.width="50%", fig.cap="Figure 3. Simple Example of a Regression Model"}
ggplot(df,
       aes(x = x,
           y = y)) +
  geom_point() +
  geom_smooth(method = "lm") + 
  theme_minimal()
```

```{r data-vs-predictions-comparison, out.width="50%", fig.cap="Figure 4. Training data and model output compared"}

@@

```{r plot-shallow-neural-network, out.width="50%", fig.cap="Figure 5. Simple Example of a Shallow Neural Network"}
knitr::include_graphics("images/nn1.png")
```

@@

```{r data-vs-predictions-comparison-2, out.width="50%", fig.cap="Figure 6. Training data and model outputs compared: linear regression and shallow neural network"}
ggplot() + 
  geom_point(data = df_pred_all,
  aes(x = x,
      y = y,
      color = type,
      shape = type,
      size = type)) +
  scale_color_manual(values = c("training data" = "black",
                                "regression output" = "blue",
                                "nn1 output" = "chocolate1",
                                "nn2 output" = "brown")) +
  scale_shape_manual(values = c("training data" = 16,
                                "regression output" = 17,
                                "nn1 output" = 18,
                                "nn2 output" = 19)) +
  scale_size_manual(values = c("training data" = 1,
                                "regression output" = 2,
                                "nn1 output" = 2,
                                "nn2 output" = 2)) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

@@

```{r plot-deeper-neural-network, out.width="50%", fig.cap="Figure 7. Example of a Deeper Neural Network"}
knitr::include_graphics("images/nn2.png")
```

@@

```{r data-vs-predictions-comparison-3, out.width="50%", fig.cap="Figure 8. Training data and model outputs compared: shallow neural network and deeper neural network"}
ggplot() + 
  geom_point(data = df_pred_all |>
               filter(type != "regression output"),
  aes(x = x,
      y = y,
      color = type,
      shape = type,
      size = type)) +
  scale_color_manual(values = c("training data" = "black",
                                "regression output" = "blue",
                                "nn1 output" = "chocolate1",
                                "nn2 output" = "brown")) +
  scale_shape_manual(values = c("training data" = 16,
                                "regression output" = 17,
                                "nn1 output" = 18,
                                "nn2 output" = 19)) +
  scale_size_manual(values = c("training data" = 1,
                                "regression output" = 2,
                                "nn1 output" = 2,
                                "nn2 output" = 2)) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

@@

```{r mac-library-econometrics, out.width="50%", fig.cap="Figure 9. Result of a search for online 'econometrics' resources in McMaster's Library catalog"}
knitr::include_graphics("images/mac-library-econometrics.png")
```

```{r mac-library-cell-biology, out.width="50%", fig.cap="Figure 10. Result of a search for online 'cell biology' resources in McMaster's Library catalog"}
knitr::include_graphics("images/mac-library-cell-biology.png")
```

```{r mac-library-euphemism, out.width="50%", fig.cap="Figure 11. Result of a search for online 'euphemism in history' resources in McMaster's Library catalogue"}
knitr::include_graphics("images/mac-library-euphemism.png")
```

@@

```{r flickr-potato, out.width="50%", fig.cap="Figure 12. Data analysis using a chatbot? Maybe not such a good idea"}
knitr::include_graphics("images/potato.png")
```

@@

```{r puzzle-web-search, eval=FALSE, out.width="50%", fig.cap="\\label{fig:puzzle-web-search}Figure 15. A search using Duck Duck Go"}
ggdraw() + 
  draw_image(image = "images/logical-puzzle-web-search.png")
```

```{r puzzle-chatgpt, out.width="50%", fig.cap="\\label{fig:puzzle-chatgpt}Figure 16. ChatGPT-4 makes the goat swim"}
puzzle_plot <- ggplot() +
  labs(caption = "https://assemblag.es/@hugo/112450401657084750") +
  theme_void()

ggdraw() + 
  draw_plot(puzzle_plot) +
  draw_image(image = "images/logical-puzzle-chat-gpt-4.png")

@@

```{r google-reddit, out.width="50%", fig.cap="Figure 13. Google's AI Overview result was traced back to a post by redditor Fucksmith"}
knitr::include_graphics("images/google-cheese-glue.png")
```

@@

```{r google-the-onion, out.width="50%", fig.cap="Figure 14. Google's AI Overview result was traced back to an article in The Onion"}
knitr::include_graphics("images/google-the-rock.jpg")
```

@@

```{r data-vs-prediction-variance}
df_pred_all |> 
  group_by(type) |> 
  summarize(variance = var(y)) |>
  transmute(Model = factor(type, 
                       levels = c("training data", "regression output", "nn1 output", "nn2 output"),
                       labels = c("Training data", "Linear regression", "Neural network 1", "Neural network 2"),
                       ordered = TRUE),
         Variance = round(variance, 2)) |>
  kable(caption = "\\ref{tab:data-vs-prediction-variance}Table 1. Variance reduction in model outputs")
```

The implication should be clear: training models with the output of models is a recipe for increasingly bland outputs. We may be witnessing _Peak ChatGPT_, and future versions of AIs, instead of improving, may begin to regress to duller versions of their own previous outputs. And this is before we even consider the possibility of AIs being deliberately "poisoned". As backlash against a data rush that is perceived as exploitative and abusive grows, new tools are being developed to generate data that are manipulated in ways that are too subtle to be detectable by a human, but that are deleterious to the training process of a GenAI [@salman_raising_2023]. One such tool is @shan_nightshade_2024's _Nightshade_, an algorithm that works to corrupt prompts used in text-to-image generation, and that notably bleeds through to poison related concepts. Another tool is @xu_shadowcast_2024's _Shadowcast_, an algorithm that poisons samples to trick an AI into misidentifying class labels, and that also can perform persuasion attacks, leveraging image-to-text generation capabilities to craft nonsensical narratives.

@@

@@

As I argue above, AIs do not do much for students or instructors, other than distract them from learning and developing genuinely transferable skills.

But AIs must do something for someone, otherwise why would they be everywhere? What do AIs do for the likes of OpenAI, Microsoft, and Google? So far, the breakneck race to develop AI has soaked immense amounts of venture capital and other speculative money, in addition to incalculable resources by many other entities, such as those spent by universities to support an AI-dedicated committee^[Some observers have even wondered whether the AI fever is not actually becoming detrimental to innovation [@siegel_ai_2023; @vinsel_dont_2023].]. Between 2012 and 2020, venture capital investments in AI grew twenty five-fold, from \$3 billion (USD) to \$75 billion (USD)^[For perspective, Ontario's Ministry of Colleges and Universities, whose mission includes supporting "research and innovation to help the province compete and thrive in the global economy", spent $163,482,281 (CAD) on its research program in 2021-2022 [@ontario_ministry_of_colleges_and_universities_expenditure_2023], a mere blip (~0.0017\%) that barely registers compared to venture capital investments in AI in 2020 alone.], much of it poured into autonomous vehicles and related mobility tech, followed by healthcare, biotechnologies and drugs, with business processes and support services in third place [@tricot_venture_2021].

@@

@berber_peter_2024, for instance, reports that the AI industry spent \$50 billion (USD) in 2023 on chips alone, while bringing in only $3 billion in revenue. @patel_inference_2023 estimated that incorporating ChatGPT-like LLMs into search would lead to a loss of \$30 billion (USD) in profit for Google. At a guess, the industry is running at a loss in almost every respect: in the same report, @patel_inference_2023 calculated the cost of running a query in ChatGPT at about \$0.36--so it would take just about 57 queries per user with a "Plus" subscription ([\$20/month](https://openai.com/chatgpt/pricing/)) for OpenAI to begin losing money on that one customer. And a large number of users are not even paying subscribers: based on revenue reports [see @perez_chatgpts_2024] there were at most 210,000 subscribers to the "Plus" plan in 2024, out of a user base estimated at tens of millions [@tong_exclusive_2023]. The profitability of AI would look even worse if not for the fact that the cost of labor is suppressed by outsourcing the most intensive, disturbing, degrading, and damaging aspects of training AIs to the Global South [@nix_analysis_2024]. The fact that the industry is not profitable has not prevented start-ups from hurriedly pivoting to GenAI, in a shift  of focus to this much hyped technology from whatever they were doing in the first place^[As if any other signs  were needed of the scammy environment in which GenAI tech thrives, @berber_peter_2024 also reports how Cognition, a startup backed by Peter Thiel, pivoted from cryptocurrencyto AI, coincidentally as cryptocurrency was going sour after the frauds at FTX were revealed. Barely disguised grift has grown abundant in the vicinity of AI at almost every level [@cook_5_2024].].

Venture capital is not known for its charitable aims, though, and at some point investors will want their money back with a heap of profit on top [@sriram_ai_2023]. Who will pay to make investors whole? And how much those who end up paying will need to pay? Or, in the worst case, _have_ to pay, if their skills become inexorably tied to this one technology? Deskilling, both of individuals and of organizations, appears to be one of the few pathways to AI profitability^[And, with profitability becoming a more distant possibility, there is always the option for large investors to cosmically inflate the hype [@salmon_what_2024] to convince bigger suckers to give them money [@reuters_openais_2024] and/or to fleece retail investors [@temkin_vcs_2024].]: once people have failed to learn a skill, and therefore become dependent on the tool, it becomes possible for the maker of the tool to turn into a digital rentier [@sadowski_internet_2020]. From this perspective, what is on offer by the AI industry is SaaS--Skill as a Service^[Universities, rather than being innovators, have increasingly become renters in the land of the platform economy [@komljenovic_rise_2021].].

@@

The key here is "labeled examples". What are those labels? Many machine algorithms do not learn by themselves...they learn from humans, and need what is called "labelled" data: a images that humans have labelled as "dog", "cats", etc. Or bits of text that have been labelled as...well...questionable. Labelling data is extremely time consuming and labor intensive. What's a corporation to do?

@@

@@

But leaving aside the question of a potential "unadmitted political agenda" by an ultra-wealthy elite, there is the much more immediate problem that training data, much like technology, are not politically neutral. Several processes have been identified that contribute to reproduce malignant biases, including that of _platformed racism_ [@matamoros-fernandez_platformed_2017]. Platformed racism results when the functionalities of platforms  (e.g., liking and sharing) are exploited to create and spread racist ideology, often in a grass-roots fashion, but more ominously with the assist of bad-faith actors^[Online Russian disinformation has already had a wide reach [@treyger_russian_2022] and furthermore is known to operate by seeding distrust between ethnic communities [@svedkauskas_analysis_2020]. A U.S. Department of State report notes that Russian intelligence is known to exploit "all kinds of separatism and ethnic, social and racial conflicts, actively supporting all dissident movements--extremist, racist, and sectarian groups" to destabilize internal U.S. politics [@us_department_of_state_gec_2024, p. 51].]. Given how cagey corporations are about the data they use to train their models, it is anyone's guess how much dis- and misinformation has been passed to their AIs^[This is part of the Unfathomable Data problem discussed by @bender_dangers_2021. In this way, AIs, rather than contributing to online safety, may end up amplifying the legacy of Russian interference [@svedkauskas_analysis_2020].]. The second aspect of platformed racism is the implicit endorsement of racist ideology through vague standards and arbitrary moderation practices^[According to the Center for Countering Digital Hate, platforms like facebook, Instagram, Tiktok, Twitter, and YouTube, failed to act 84% of the time on hundreds of posts that had millions of views, with facebook being the worst performer [@center_for_countering_digital_hate_failure_2021].] that exclude some views while allowing others to thrive [@myers_west_censored_2018; @hawkins_race_2023].

Search engines were already known to reinforce racism [@noble_algorithms_2018]; given the above, it really is too much to expect that AI-augmented search engines will do any better^[A report by the Center for Countering Digital Hate found that Google's Bard generated wrong and/or hateful outputs in 78 out of 100 tests [@center_for_countering_digital_hate_googles_2023]. In addition to being misleading to the unadvised, AI is also a fantastic tool ripe for abuse by those who are advised _and_ have an agenda. Google's own researchers have come to the inevitable conclusion that it is easier to produce disinformation with AI, than it is to use AI to combat it [@maiberg__google_2024].]. In another wretched incident, Google's AI Overview stated that Barack Obama was the first Muslim president of the US, a regrettable piece of misinformation that fails to surprise given the years-long Russian-assisted right-wing propaganda campaign to otherize Mr. Obama [@tenbarge_glue_2024]. Mishaps like this, and others, are dismissed by Google as attempts "to trip up the technology with uncommon questions" [@tenbarge_glue_2024]. This dismissal, arrogant as it is, only reinforces the notion that AIs are only ever useful when [regressing to the mean](#really-what-do-they-do). Responses like fiddling with the model (fine tuning it) to provide more palatable outputs^[Like in the case of [Google's response](https://blog.google/products/search/ai-overviews-update-may-2024/) to the poor quality of its recently launched AI overviews.] are not reassuring. Who is to say what a handful of unaccountable corporations will find unpalatable tomorrow given their insatiable quest for profits, let alone in the face of demands by powerful authoritarian regimes^[@vandehei_behind_2024 write in Axios about an "ominous scarcity" of inputs that threatens America's AI superiority over China. These authors, whose one true expertise is access journalism, never stop to consider why an authoritarian regime might want to have powerful pattern recognition machines. To compound things, they neglect to disclose the stake that Axios, the website they founded, has on AI. VandeHei has not been timid to express himself otherwise, and is on record stating that AI "will eviscerate the weak, the ordinary, the unprepared in media" [@robertson_axios_2024]. Who should be concerned about maintaining an edge?]?

```{r google-obama, out.width="50%", fig.cap="\\label{fig:google-ai-overview-of-obama}Figure 17. Google's AI Overview of Muslim US presidents"}
google_plot <- ggplot() +
  labs(caption = "Source: https://www.nbcnews.com/tech/tech-news/google-ai-im-feeling-depressed-cheese-not-sticking-to-pizza-error-rcna153301") +
  theme_void()

ggdraw() + 
  draw_plot(google_plot) +
  draw_image(image = "images/google-obama.png")
```

@@

Neural networks and machine learning are not new, and they have their uses. But those uses have been drown, and to some extent discredited, by the current hype around AI.

@morrone_signals_2024 Carefully manufactured and well funded hype.

ChatGPT was a colossal feat of marketing [@cowen_chatgpt_2023], but much coverage of AI in the media has been stretched beyond the breaking point of credulity [@burneko_if_2024]^[Bob Woodward said "journalism is not stenography" [@hiskes_woodward_2007], but it is difficult to see what else writing like Roose's [-@roose_is_2024] could be called, besides hagiography.]. There are now calls to press pause on the hype [@angwin_press_2024] but these are still drown by the uncritical coverage.

> An end to “batch” teaching. Instead of working with cohorts of learners in grades or classes, individual learning agendas are developed based on an assessment of the skills, competencies and capabilities the learner already has as well as ongoing conversations about the purpose of the learning. Where does the student want the learning to take them? AI systems, coupled with coaching, mentoring and some instruction and peer support networks then ensure the learner is supported on the carefully mapped learning journey. Teaching can still take place: Students challenged with a set of related issues can be brought together for a boot camp or focused period of instruction before returning to their individualized study activities.

@@

> Greater accessibility. Higher education systems are organized around blocks of time: semesters or terms. There is no longer good reason for this. Students should be able to begin their learning at any time and be assessed when they are ready. A four-year undergraduate degree could easily be completed in 16 months by a diligent student, for example, or in even less time by someone arriving in the program with substantial life and work experience. AI systems can also leverage technologies to ensure the resources and supports for learners reflect who they are (culture, language, experience and skill) and what challenges they face (disabilities, learning styles, etc.).

>>>>

---
title: "An AI is haunting campus"
subtitle: "Some considerations about the use of AI for teaching and learning in post-secondary education"
author: Antonio Páez (McMaster University)
date: "`r Sys.Date()`"
format:
    pdf: 
      fontsize: 7pt
    html: 
      self-contained: true
      grid: 
        margin-width: 350px
execute: 
  echo: fenced
  freeze: auto
reference-location: margin
#citation-location: margin
bibliography:
  - "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
---

@@

::: {.column-margin}
Karl Marx wrote about a "spectre...haunting Europe--the spectre of communism"; GenAI looks awfully like the spectre of unregulated surveillance capitalism.
:::

@@

Note: _This essay was written as an open research project, and it can be found in the following repository:_ https://github.com/paezha/University-AI-Panel

## Background

@@

# The university's mission {#the-mission}

@@

```{r ai-writing-code, out.width="70%", fig.cap="\\label{fig:ai-bug-fix}An example of Atlassian AI writing code"}
knitr::include_graphics(glue::glue(here::here(), "/images/ai-bug-fix.png"))
```

@@

```{r ai-litter-box, out.width="70%", fig.cap="\\label{fig:ai-bug-fix}Revolutionizing pet health"}
knitr::include_graphics(glue::glue(here::here(), "/images/ai-litter-box.png"))
```

@@

\newpage

@@

```{r plot-simulated-data, message=FALSE, out.width="70%", fig.cap="Simple Example of a Regression Model"}
ggplot(df,
       aes(x = x,
           y = y)) +
  geom_point() +
  geom_smooth(method = "lm") + 
  theme_minimal()
```

```{r data-vs-predictions-comparison, out.width="70%", fig.cap="Training data and model output compared"}

@@

```{r plot-shallow-neural-network, out.width="70%", fig.cap="Simple Example of a Shallow Neural Network"}
knitr::include_graphics(glue::glue(here::here(), "/images/nn1.png"))
```

@@

```{r data-vs-predictions-comparison-2, out.width="70%", fig.cap="Training data and model outputs compared: linear regression and shallow neural network"}
ggplot() + 
  geom_point(data = df_pred_all,
  aes(x = x,
      y = y,
      color = type,
      shape = type,
      size = type)) +
  scale_color_manual(values = c("training data" = "black",
                                "regression output" = "blue",
                                "nn1 output" = "chocolate1",
                                "nn2 output" = "brown")) +
  scale_shape_manual(values = c("training data" = 16,
                                "regression output" = 17,
                                "nn1 output" = 18,
                                "nn2 output" = 19)) +
  scale_size_manual(values = c("training data" = 1,
                                "regression output" = 2,
                                "nn1 output" = 2,
                                "nn2 output" = 2)) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

@@

```{r plot-deeper-neural-network, out.width="70%", fig.cap="Example of a Deeper Neural Network"}
knitr::include_graphics(glue::glue(here::here(), "/images/nn2.png"))
```

@@

```{r data-vs-predictions-comparison-3, out.width="70%", fig.cap="Training data and model outputs compared: shallow neural network and deeper neural network"}
ggplot() + 
  geom_point(data = df_pred_all |>
               filter(type != "regression output"),
  aes(x = x,
      y = y,
      color = type,
      shape = type,
      size = type)) +
  scale_color_manual(values = c("training data" = "black",
                                "regression output" = "blue",
                                "nn1 output" = "chocolate1",
                                "nn2 output" = "brown")) +
  scale_shape_manual(values = c("training data" = 16,
                                "regression output" = 17,
                                "nn1 output" = 18,
                                "nn2 output" = 19)) +
  scale_size_manual(values = c("training data" = 1,
                                "regression output" = 2,
                                "nn1 output" = 2,
                                "nn2 output" = 2)) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

@@

```{r mac-library-econometrics, out.width="70%", fig.cap="Result of a search for online 'econometrics' resources in McMaster's Library catalog"}
knitr::include_graphics(glue::glue(here::here(), "/images/mac-library-econometrics.png"))
```

```{r mac-library-cell-biology, out.width="70%", fig.cap="Result of a search for online 'cell biology' resources in McMaster's Library catalog"}
knitr::include_graphics(glue::glue(here::here(), "/images/mac-library-cell-biology.png"))
```

```{r mac-library-euphemism, out.width="70%", fig.cap="Result of a search for online 'euphemism in history' resources in McMaster's Library catalogue"}
knitr::include_graphics(glue::glue(here::here(), "/images/mac-library-euphemism.png"))
```

@@

```{r flickr-potato, out.width="70%", fig.cap="Data analysis using a chatbot? Maybe not such a good idea"}
knitr::include_graphics(glue::glue(here::here(), "/images/potato.png"))
```

@@

```{r puzzle-web-search, eval=FALSE, out.width="70%", fig.cap="A search using Duck Duck Go"}
ggdraw() + 
  draw_image(image = glue::glue(here::here(), "/images/logical-puzzle-web-search.png"))
```

```{r puzzle-chatgpt, out.width="70%", fig.cap="ChatGPT-4 makes the goat swim"}
puzzle_plot <- ggplot() +
  labs(caption = "https://assemblag.es/@hugo/112450401657084750") +
  theme_void() +
  theme(text = element_text(size = 8))

ggdraw() + 
  draw_plot(puzzle_plot) +
  draw_image(image = glue::glue(here::here(), "/images/logical-puzzle-chat-gpt-4.png"))

@@

```{r google-reddit, out.width="70%", fig.cap="Google's AI Overview result was traced back to a post by redditor Fucksmith"}
knitr::include_graphics(glue::glue(here::here(), "/images/google-cheese-glue.png"))
```

@@

```{r google-the-onion, out.width="70%", fig.cap="Google's AI Overview result was traced back to an article in The Onion"}
knitr::include_graphics(glue::glue(here::here(), "/images/google-the-rock.jpg"))
```

\newpage

@@

```{r data-vs-prediction-variance}
df_pred_all |> 
  group_by(type) |> 
  summarize(variance = var(y)) |>
  transmute(Model = factor(type, 
                       levels = c("training data", "regression output", "nn1 output", "nn2 output"),
                       labels = c("Training data", "Linear regression", "Neural network 1", "Neural network 2"),
                       ordered = TRUE),
         Variance = round(variance, 2)) |>
  kable(caption = "Variance reduction in model outputs")
```

The implication should be clear: training models with the output of models is a recipe for increasingly bland outputs. We may be witnessing _Peak ChatGPT_, and future versions of AIs, instead of improving, may begin to regress to duller versions of their own previous outputs. And this is before we even consider the possibility of AIs being deliberately "poisoned". As backlash against a data rush that is perceived as exploitative and abusive grows, new tools are being developed to generate data that are manipulated in ways that are too subtle to be detectable by a human, but that are deleterious to the training process of a GenAI [@salman_raising_2023]. One such tool is _PhotoGuard_, developed by a team at MIT to send signals to an AI that prevents it from manipulating images [@heikkila_this_2023]. Other examples include @shan_nightshade_2024's _Nightshade_, an algorithm that works to corrupt prompts used in text-to-image generation, and that notably bleeds through to poison related concepts. And more recently, @xu_shadowcast_2024 presented _Shadowcast_, an algorithm that poisons samples to trick an AI into misidentifying class labels, and that also can perform persuasion attacks, leveraging image-to-text generation capabilities to craft nonsensical narratives.

@@

## What do GenAIs do for the Province?

As I argue above, AIs do not seem to do much for students or instructors, other than distract them from learning and developing genuinely transferable skills. On the other hand, it is hard not to suspect that the provincial government is hoping to use GenAIs to further reduce the need to fund higher education. But AIs must do something for someone, otherwise why would they be everywhere? For Ontario, the main patron and effectively the boss of McMaster, it provides a rationale (an excuse even) to force the institution to find "efficiencies". Tom Goldstein, a computer science professor at the University of Maryland, thinks that the lure of GenAI will be hard to resist for many, irrespective of their flaws...after all, "[e]ven though [GenAI is] expensive...it's still far less expensive than human labor" [@oremus_chatgpt_2023]. Without adjudicating the merits of the argument that human labor is expensive^[When the alternative is substandard quality of life for many of our fellow citizens, and/or higher unemployment.], it is hard to argue against the lure of GenAI for governments that do not seem to believe in higher education.

@@

But neither the province nor the university are developing the AIs that they hope will make them cheaper to operate. So what do AIs do for the the entities that develop them, namely OpenAI et al.? So far, the breakneck race to develop AI has soaked immense amounts of venture capital and other speculative money, in addition to incalculable resources by many other entities, such as those spent by universities to support an AI-dedicated committee^[Some observers have even wondered whether the AI fever has not actually become detrimental to innovation [@siegel_ai_2023; @vinsel_dont_2023].]. Between 2012 and 2020, venture capital investments in AI grew twenty five-fold, from \$3 billion (USD) to \$75 billion (USD)^[For perspective, Ontario's Ministry of Colleges and Universities, whose mission includes supporting "research and innovation to help the province compete and thrive in the global economy", spent $163,482,281 (CAD) on its research program in 2021-2022 [@ontario_ministry_of_colleges_and_universities_expenditure_2023], a mere blip (~0.0017\%) that barely registers compared to venture capital investments in AI in 2020 alone.], much of it poured into autonomous vehicles and related mobility tech, followed by healthcare, biotechnologies and drugs, with business processes and support services in third place [@tricot_venture_2021].

@@

@berber_peter_2024, for instance, reports that the AI industry spent \$50 billion (USD) in 2023 on chips alone, while bringing in only $3 billion in revenue. @patel_inference_2023 estimated that incorporating ChatGPT-like LLMs into search would lead to a loss of \$30 billion (USD) in profit for Google. At a guess, the industry is running at a loss in almost every respect: in the same report, @patel_inference_2023 calculated the cost of running a query in ChatGPT at about \$0.36--so it would take just about 57 monthly queries per user with a "Plus" subscription ([\$20/month](https://openai.com/chatgpt/pricing/)) for OpenAI to begin losing money on that one customer. And a large number of users are not even paying subscribers: based on revenue reports [see @perez_chatgpts_2024] there were at most 210,000 subscribers to the "Plus" plan in 2024, out of a user base estimated at tens of millions [@tong_exclusive_2023]. The profitability of AI would look even worse if not for the fact that the cost of labor is suppressed by outsourcing the most intensive, disturbing, degrading, and damaging aspects of training AIs to the Global South [@nix_analysis_2024]. The fact that the industry is not profitable has not prevented start-ups from hurriedly pivoting to GenAI, in a shift  of focus to this much hyped technology from whatever they were doing in the first place^[As if any other signs  were needed of the scammy environment in which GenAI tech thrives, @berber_peter_2024 also reports how Cognition, a startup backed by Peter Thiel, pivoted from cryptocurrency to AI, coincidentally as cryptocurrency was going sour after the frauds at FTX were revealed. Barely disguised grift has grown abundant in the vicinity of AI at almost every level [@cook_5_2024].].

Venture capital is not known for its charitable aims, though, and at some point investors will want their money back with a heap of profit on top [@sriram_ai_2023]. Who will pay to make investors whole? And how much those who end up paying will need to pay? Or, in the worst case, _have_ to pay, if their skills become inexorably tied to this one technology? Deskilling, both of individuals and of organizations, appears to be one of the few pathways to AI profitability^[And, with profitability increasingly looking like a distant possibility, there is always the option for large investors to cosmically inflate the hype [@salmon_what_2024] in order to convince bigger suckers to give them money [@reuters_openais_2024] and/or to fleece retail investors [@temkin_vcs_2024].]: once people have failed to learn a skill, and therefore become dependent on the tool, it becomes possible for the maker of the tool to turn into a digital rentier [@sadowski_internet_2020]. From this perspective, what is on offer by the AI industry is SaaS--Skill as a Service^[Universities, rather than being innovators, have increasingly become renters in the land of the platform economy [@komljenovic_rise_2021].].

@@

The key here is "labelled examples". What are those labels? Many machine algorithms do not learn by themselves...they learn from humans, and need what is called "labelled" data: a images that humans have labelled as "dog", "cats", etc. Or bits of text that have been labelled as...well...questionable. Labeling data is extremely time consuming and labor intensive. What's a corporation to do?

@@

```{r google-obama, out.width="90%", fig.cap="\\label{fig:google-ai-overview-of-obama}Google's AI Overview of Muslim US presidents"}
google_plot <- ggplot() +
  labs(caption = "Source: https://www.nbcnews.com/tech/tech-news/google-ai-im-feeling-depressed-cheese-not-sticking-to-pizza-error-rcna153301") +
  theme_void() +
  theme(text = element_text(size = 8))

ggdraw() + 
  draw_plot(google_plot) +
  draw_image(image = glue::glue(here::here(), "/images/google-obama.png"))
```

@@

But leaving aside the question of a potential "unadmitted political agenda" by an ultra-wealthy elite, there is the much more immediate problem that training data, much like technology, are not politically neutral. Several processes have been identified that contribute to reproduce malignant biases, including that of _platformed racism_ [@matamoros-fernandez_platformed_2017]. Platformed racism results when the functionalities of platforms  (e.g., liking and sharing) are exploited to create and spread racist ideology, often in a grass-roots fashion, but more ominously with the assist of bad-faith actors^[Online Russian disinformation has already had a wide reach [@treyger_russian_2022] and furthermore is known to operate by seeding distrust between ethnic communities [@svedkauskas_analysis_2020]. A U.S. Department of State report notes that Russian intelligence is known to exploit "all kinds of separatism and ethnic, social and racial conflicts, actively supporting all dissident movements--extremist, racist, and sectarian groups" to destabilize internal U.S. politics [@us_department_of_state_gec_2024, p. 51].]. Given how cagey corporations are about the data they use to train their models, it is anyone's guess how much dis- and misinformation has been passed to their AIs^[This is part of the Unfathomable Data problem discussed by @bender_dangers_2021. In this way, AIs, rather than contributing to online safety, may end up amplifying the legacy of Russian interference [@svedkauskas_analysis_2020].]. The second aspect of platformed racism is the implicit endorsement of racist ideology through vague standards and arbitrary moderation practices^[According to the Center for Countering Digital Hate, platforms like facebook, Instagram, Tiktok, Twitter, and YouTube, failed to act 84% of the time on hundreds of posts that had millions of views, with facebook being the worst performer [@center_for_countering_digital_hate_failure_2021]. This is not a bug of the system, but rather a feature, as Meta's supression of information about Gaza in 2024 shows [@paul_former_2024].] that exclude some views while allowing others to thrive [@myers_west_censored_2018; @hawkins_race_2023].

Search engines were already known to reinforce racism [@noble_algorithms_2018]; given the above, it really is too much to expect that AI-augmented search engines will do any better^[A report by the Center for Countering Digital Hate found that Google's Bard generated wrong and/or hateful outputs in 78 out of 100 tests [@center_for_countering_digital_hate_googles_2023]. In addition to being misleading to the unadvised, AI is also a fantastic tool ripe for abuse by those who are advised _and_ have an agenda. Google's own researchers have come to the inevitable conclusion that it is easier to produce disinformation with AI, than it is to use AI to combat it [@maiberg__google_2024].]. In another wretched incident, Google's AI Overview stated that Barack Obama was the first Muslim president of the US, a regrettable piece of misinformation that fails to surprise given the years-long Russian-assisted right-wing propaganda campaign to otherize Mr. Obama [@tenbarge_glue_2024]. Mishaps like this, and others, are dismissed by Google as attempts "to trip up the technology with uncommon questions" [@tenbarge_glue_2024]. This dismissal, arrogant as it is, only reinforces the notion that AIs are only ever useful when [regressing to the mean](#really-what-do-they-do). Responses like fiddling with the model (fine tuning it) to provide more palatable outputs^[Like in the case of [Google's response](https://blog.google/products/search/ai-overviews-update-may-2024/) to the poor quality of its recently launched AI overviews.] are not reassuring.

\newpage

Who is to say what a handful of unaccountable corporations will find unpalatable tomorrow given their insatiable quest for profits, let alone in the face of demands by powerful authoritarian regimes^[@vandehei_behind_2024 write in Axios about an "ominous scarcity" of inputs that threatens America's AI superiority over China. These authors, whose one true expertise is access journalism, never stop to consider why an authoritarian regime might want to have powerful pattern recognition machines. To compound things, they neglect to disclose the stake that Axios, the website they founded, has on AI. VandeHei has not been timid to express himself otherwise, and is on record stating that AI "will eviscerate the weak, the ordinary, the unprepared in media" [@robertson_axios_2024]. Who is concerned about maintaining an edge?]?

@@

Neural networks and machine learning are nothing new, and they have their uses. But those uses have been suffocated, and to some extent discredited, by the current hype around Generative AI. This hype has been carefully manufactured [@morrone_signals_2024], and has been amplified by media tools that have been more than willing to believe in magic beans [@burneko_if_2024]^[Bob Woodward said "journalism is not stenography" [@hiskes_woodward_2007], but it is difficult to see what else writing like Roose's [-@roose_is_2024] could be called, besides hagiography.]. There are now calls to press pause on the hype [@angwin_press_2024] but they are barely audible above the reverberating echos inside the AI hype chamber.

Universities and colleges would be wise to heed the advice to press the pause button on the AI hype, as there is a real risk of them falling prey to a confidence trick. Higher education is often driven by a deeply ingrained belief in innovation (even, or perhaps particularly, for its own sake), and this impulse is compounded by the way AI adoption has been drawn as a false dilemma, thus triggering a fear of missing out. There are other pressures at work, including a decades-long history of manufactured crises in education, in Canada as elsewhere [@cizek_give_1999; @usher_state_2023].

In this essay I have tried to describe the many problems with AI as an _everything tool_. Tall in my mind are the moral and ethical concerns around an extractive and exploitative technology that promises democracy but delivers further concentration of wealth and power.

<!--
> An end to “batch” teaching. Instead of working with cohorts of learners in grades or classes, individual learning agendas are developed based on an assessment of the skills, competencies and capabilities the learner already has as well as ongoing conversations about the purpose of the learning. Where does the student want the learning to take them? AI systems, coupled with coaching, mentoring and some instruction and peer support networks then ensure the learner is supported on the carefully mapped learning journey. Teaching can still take place: Students challenged with a set of related issues can be brought together for a boot camp or focused period of instruction before returning to their individualized study activities.

@@

> Greater accessibility. Higher education systems are organized around blocks of time: semesters or terms. There is no longer good reason for this. Students should be able to begin their learning at any time and be assessed when they are ready. A four-year undergraduate degree could easily be completed in 16 months by a diligent student, for example, or in even less time by someone arriving in the program with substantial life and work experience. AI systems can also leverage technologies to ensure the resources and supports for learners reflect who they are (culture, language, experience and skill) and what challenges they face (disabilities, learning styles, etc.).
-->
